---
title: "magrittr + dplyr + tidyr"
author: "Vitor Aguiar"
date: ""
output: 
    ioslides_presentation:
      widescreen: true
      smaller: true
---

## magrittr

## magrittr {.flexbox .vcenter}

<big style="font-size:10em"> *%>%* </big>

[Ceci n'est pas une pipe](http://uploads7.wikiart.org/images/rene-magritte/the-treachery-of-images-this-is-not-a-pipe-1948%282%29.jpg)

## magrittr operators {.flexbox .vcenter}

<big style="font-size:2em"> *%>%* </big>

<big style="font-size:2em"> *%T>%* </big>

<big style="font-size:2em"> *%$%* </big>

<big style="font-size:2em"> *%<>%* </big>

## pipe usage

```{r eval=FALSE}
f(x)

# is the same as:

x %>% f()

## with two arguments:
f(x, y)

# is the same as:
x %>% f(y)

## if you don't want the input to be used as 1st argument:
f(y, x)

# is the same as:
x %>% f(y, .)

```

## {}

<big style="font-size:5em"> *dplyr* </big>
(aka plyr for data frames)

## dplyr verbs {.flexbox .vcenter}

filter

slice

group_by

arrange

select

summarise

mutate

## {}

We'll use the [gapminder](https://www.youtube.com/watch?v=jbkSRLYSojo) data for a demo

```{r}
library(gapminder)

str(gapminder)
```

## {}

```{r message=FALSE}
library(dplyr)
```

```{r, comment=NA}
gapminder <- tbl_df(gapminder)

gapminder
```

## count

```{r, comment=NA}
gapminder %>% count(country)
```

## filter and select

```{r, comment=NA}
gapminder %>%
  filter(country == "Brazil", year >= 1980) %>%
  select(year, pop, lifeExp, gdpPercap)
```

## arrange

```{r, comment=NA}
gapminder %>%
  arrange(lifeExp) %>%
  slice(1:10)
```

## arrange by multiple variables

```{r, comment=NA}
gapminder %>%
  arrange(desc(pop), gdpPercap, lifeExp)
```

## group_by + summarise

```{r, comment=NA}
gapminder %>%
  group_by(continent) %>%
  summarise(avg_gdp = mean(gdpPercap))
```

## group_by + summarise: multiple grouping variables

```{r, comment=NA}
gapminder %>%
  group_by(continent, year) %>%
  summarise(avg_gdp = mean(gdpPercap))
```

## summarise: multiple variables

```{r, comment=NA}
gapminder %>%
  group_by(continent) %>%
  summarise_each(funs(mean), lifeExp, gdpPercap, pop)
```

## summarise: mutiple functions & multiple variables

```{r, comment=NA}
gapminder %>%
  group_by(continent) %>%
  summarise_each(funs(mean, sd), lifeExp, gdpPercap)
```

## mutate

```{r, comment=NA}
gapminder %>% mutate(total_gdp = gdpPercap * pop)
```

## mutate_each

```{r, comment=NA}
gapminder %>% 
  mutate_each(funs(min_rank), lifeExp, gdpPercap, pop)
```

## data %>% plot

```{r, fig.width=6, fig.height=3, fig.align="center"}
library(ggplot2)

gapminder %>%
  mutate(total_gdp = gdpPercap * pop) %>% 
  group_by(continent, year) %>%
  summarise(total_gdp = sum(total_gdp)) %>%
  ggplot(data = ., aes(x = year, y = total_gdp, color = continent)) + geom_line()
```

## {}

<big style="font-size:5em"> *tidyr* </big>

## Plant height data measured in 3 different months

```{r echo=FALSE, comment=NA}
plant_height <- data_frame(individual = paste(sample(1:20), 
                                              c(rep("SP", 10), rep("MG", 10)),
                                              sep = "_"),
                         "2015/01/01" = rnorm(20, 15, sd = 1),
                         "2015/02/01" = rnorm(20, 16, sd = 1.2),
                         "2015/03/01" = rnorm(20, 21, sd = 1.5))

plant_height
```

## tidyr verbs {.flexbox .vcenter}

gather 

spread

separate

unite

extract

## gather: 'long' or 'narrow' format

```{r message=FALSE, comment=NA}
library(magrittr)
library(tidyr)

plant_height %<>% gather(key=date, value=height, 2:4)

plant_height
```

## spread

```{r, comment=NA}
plant_height %>% spread(date, height)
```

## separate

```{r, comment=NA}
plant_height %<>% 
  separate(individual, into = c("id_number", "state")) %>%
  mutate(id_number = as.numeric(id_number)) %>%
  arrange(id_number, state)

plant_height
```

## extract

```{r, comment=NA}
plant_height %<>% 
  extract(date, c("year", "month", "day"), "(\\d+)/(\\d+)/(\\d+)")

plant_height
```

## unite 

```{r, comment=NA}
plant_height %>% unite(date, year:day, sep = "/")
```

## transpose data frame

```{r, comment=NA}
plant_height %>% spread(id_number, height)
```

## Data wrangling cheat sheet

[Data manipulation cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
